---
title: "Project1_EDA_draft"
author: "Erika Pham"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
```{r init, include = FALSE}
# loading in libraries
library(ezids)
library(tidyverse)
library(readr)
library(ggplot2)
library(stringr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
options(scientific = T, digits = 3)
```


```{r, include = FALSE}
# loading in dataset
diabetes <- read.csv("diabetes_binary_5050split_health_indicators_BRFSS2015.csv")
str(diabetes)

# renaming some longer variables
diabetes <- diabetes %>%
    rename("diabetes"=Diabetes_binary)
diabetes <- diabetes %>%
  rename("heart_dis" = HeartDiseaseorAttack)
diabetes <- diabetes %>%
  rename("hvy_alc" = HvyAlcoholConsump)
```



```{r, include = FALSE}
diabts1 <- subset(diabetes, diabetes == 1)
diabts0 <- subset(diabetes, diabetes == 0)
```
### Bar charts for each distribution of health factors among people with diabetes
**For high blood pressure**

```{r}
library(ggplot2)
highbp1 <- ggplot(data = diabetes, aes(x = factor(HighBP), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "High blood pressure distribution", x = "have high blood pressure") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=c("lightskyblue","#E69234"))
    
highbp1
```

**For high cholesterol**
**1: has high BP, 0: does not have high BP**

``` {r}
highchol1 <- ggplot(data = diabetes, aes(x = factor(HighChol), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "High cholesterol level distribution", x = "have high cholesterol") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="Set1")
    
highchol1

```
**For CholCheck (if the person has checked for cholesterol level within the past 5 years)**

```{r}
cholcheck1 <- ggplot(data = diabetes, aes(x = factor(CholCheck), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Cholesterol-check within 5 years distribution", x = "checked") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="Accent")
    
cholcheck1

```

**For BMI levels**

```{r}
bmi1 <- ggplot(data = diabetes, aes(x = BMI, fill=factor(diabetes))) +
      geom_boxplot(color = "black")+
      labs(title = "Distribution of BMI", x = "BMI") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5),
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank())+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="Paired")
bmi1
```

**For Smoker**
```{r}
smoker1 <- ggplot(data = diabetes, aes(x = factor(Smoker), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Distribution of smokers", x = "smoking") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="PuRd")
    
smoker1
```

**For Stroke**

```{r}
stroke1 <- ggplot(data = diabetes, aes(x = factor(Stroke), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Stroke distribution", x = "have had a stroke") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="YlGn")
    
stroke1
```

**For Heart Disease/Attack**

```{r}
heart_dis1 <- ggplot(data = diabetes, aes(x = factor(heart_dis), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Heart disease/attack distribution", x = "have heart disease/attack") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="Pastel1")
    
heart_dis1
```

**For PhysActivity (if they've had any physical activity the past 30 days)**

```{r}
phys1 <- ggplot(data = diabetes, aes(x = factor(PhysActivity), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Physical activity distribution", x = "physical activity within 30 days") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="Spectral")
    
phys1
```
**For Fruits (daily consumption of fruits)**

```{r}
fruits1 <- ggplot(data = diabetes, aes(x = factor(Fruits), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Daily fruits consumption distribution", x = "had fruits") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="PuOr")
    
fruits1
```

**For Veggies**

```{r}
veg1 <- ggplot(data = diabetes, aes(x = factor(Veggies), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Daily consumption of vegetables", x = "had vegetables") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="Purples")
    
veg1
```

**For HvyAlcohol (adult men >= 14 drinks/wk, adult women >= 7 drinks/wk)**
```{r}
library(ggplot2)
hvyalc1 <- ggplot(data = diabetes, aes(x = factor(hvy_alc), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "High alcohol consumption distribution", x = "high-alcohol consumption") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_brewer(palette="Set3")
    
hvyalc1
```


**For AnyHealthcare**
```{r}
library(ggplot2)
any_health1 <- ggplot(data = diabetes, aes(x = factor(AnyHealthcare), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Health insurance availability", x = "have insurance") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("mistyrose","skyblue"))
any_health1
```


**For NoDocbcCost (needed to see a doctor but did not due to cost in the past 12 months)**

```{r}
library(ggplot2)
nodoc1 <- ggplot(data = diabetes, aes(x = factor(NoDocbcCost), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Refusal of doctor visits due to cost", x = "did not see doctor") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("#E78321","#927999"))
nodoc1
```

**For GenHlth**


```{r}
genhlth1 <- ggplot(data = diabetes, aes(x = GenHlth, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of self-rated general health level (1 to 5)", x = "general health level") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("lightsalmon","paleturquoise"))
genhlth1
```

**For MentHlth (no. days of poor mental health, 1-30 days)**

```{r}
mentalh1 <- ggplot(data = diabts1, aes(x = MentHlth)) +
      geom_histogram(binwidth = 10, width = 0.5, color="black", fill="plum3")+
      labs(title = "Distribution of poor mental health days among people with pre/diabetes", x = "no. of days") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))
mentalh1
```

```{r}
mentalh0 <- ggplot(data = diabts0, aes(x = MentHlth)) +
      geom_histogram(binwidth = 10, width = 0.5, color="black", fill="lightpink")+
      labs(title = "Distribution of poor mental health days \namong people without pre/diabetes", x = "no. of days") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))
mentalh0
```

**For DiffWalk (if participant had serious difficulty walking/climbing stairs**

```{r}
diffwalk1 <- ggplot(data = diabetes, aes(x = factor(DiffWalk), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Difficulty walking/climbing stairs distribution", x = "have difficulty walking") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("gold3","brown2"))
diffwalk1
```

**For Sex**

```{r}
sex1 <- ggplot(data = diabetes, aes(x = factor(Sex), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of sex", x = "sex (0 = female, 1 = male)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("green4","yellow4"))
sex1
```

**For Age**
```{r}
age1 <- ggplot(data = diabetes, aes(x = Age, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of age groups", x = "age groups") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("pink3","lightsteelblue"))
age1
```

**For Education**

```{r}
edu1 <- ggplot(data = diabetes, aes(x = Education, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of education levels", x = "education level") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("cornflowerblue","darkorange"))
edu1
```

**For Income**


```{r}
income1 <- ggplot(data = diabetes, aes(x = Income, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of income levels", x = "income level") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values = c("blueviolet","coral2"))
income1
```

