---
title: "DATS6101 Project 1"
author: Group 5
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
date: "2023-03-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F, message = F)
options(scientific = T, digits = 3)
```

```{r init, include = FALSE}
# loading in libraries
library(ezids)
library(tidyverse)
library(readr)
library(ggplot2)
library(stringr)
library(gridExtra)
```

```{r, include = FALSE}
# loading in dataset
diabetes <- read.csv("diabetes_binary_5050split_health_indicators_BRFSS2015.csv")
str(diabetes)

# renaming some longer variables
diabetes <- diabetes %>%
    rename("diabetes"= Diabetes_binary)
diabetes <- diabetes %>%
  rename("heart_dis" = HeartDiseaseorAttack)
diabetes <- diabetes %>%
  rename("hvy_alc" = HvyAlcoholConsump)

# storing categorical variables as factor
diabetes$diabetes <- as.factor(diabetes$diabetes)
diabetes$HighBP <- as.factor(diabetes$HighBP)
diabetes$HighChol <- as.factor(diabetes$HighChol)
diabetes$CholCheck <- as.factor(diabetes$CholCheck)
diabetes$Smoker <- as.factor(diabetes$Smoker)
diabetes$Stroke <- as.factor(diabetes$Stroke)
diabetes$NoDocbcCost <- as.factor(diabetes$NoDocbcCost)
diabetes$heart_dis <- as.factor(diabetes$heart_dis)
diabetes$PhysActivity <- as.factor(diabetes$PhysActivity)
diabetes$Fruits <- as.factor(diabetes$Fruits)
diabetes$Veggies <- as.factor(diabetes$Veggies)
diabetes$hvy_alc <- as.factor(diabetes$hvy_alc)
diabetes$DiffWalk <- as.factor(diabetes$DiffWalk)
diabetes$Sex <- as.factor(diabetes$Sex)
diabetes$GenHlth <- as.factor(diabetes$GenHlth)
diabetes$AnyHealthcare <- as.factor(diabetes$AnyHealthcare)
diabetes$Education <- as.factor(diabetes$Education)
diabetes$Income <- as.factor(diabetes$Income)

```

```{r, include = FALSE}
# colorblind color palettes

# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# To use for fills, add
  #scale_fill_manual(values=cbPalette)

# To use for line and point colors, add
  #scale_colour_manual(values=cbPalette)
```

# Examination of 

Abstract here

#### 1. Background
#### 2. Data
#### 3. EDA

```{r, include = FALSE}
# subseting the data between people who have pre/diabetes and people who do not
diabts1 <- subset(diabetes, diabetes == 1)
diabts0 <- subset(diabetes, diabetes == 0)
```

**First, we ran some summary statistics for the data set:**
```{r, comment = ""}
summary(diabetes)
```
Since most of our features are categorical, running summary statistics will not be very helpful in seeing what is going on. We decided that graphs and plots, particularly bar graphs and histograms are much more effective in portraying what the data looks like and what are the potential things we could look at. 

**Graphs for each variable/feature**

**For high blood pressure**


```{r}
library(ggplot2)
highbp1 <- ggplot(data = diabetes, aes(x = factor(HighBP), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "High blood pressure distribution", x = "have high blood pressure (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
highbp1
```

**For high cholesterol**

**1: has high BP, 0: does not have high BP**

``` {r}
highchol1 <- ggplot(data = diabetes, aes(x = factor(HighChol), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "High cholesterol level distribution", x = "have high cholesterol (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
highchol1

```

**For CholCheck (if the person has checked for cholesterol level within the past 5 years)**

```{r}
cholcheck1 <- ggplot(data = diabetes, aes(x = factor(CholCheck), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Cholesterol-check within 5 years distribution", x = "checked (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
cholcheck1

```

**For BMI levels**

```{r}
bmi <- ggplot(data = diabetes, aes(x = BMI, fill=factor(diabetes))) +
      geom_boxplot(color = "black")+
      labs(title = "Distribution of BMI", x = "BMI") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5),
            axis.text.y = element_blank(),
            axis.ticks.y = element_blank())+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
bmi
```

```{r}
bmi1 <- ggplot(data = diabts1, aes(x = BMI)) +
      geom_histogram(width = 0.5, color="black", fill="plum3")+
      labs(title = "Distribution of BMI among \npeople with pre/diabetes", x = "BMI") +
      theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
bmi1

bmi0 <- ggplot(data = diabts0, aes(x = BMI)) +
      geom_histogram(width = 0.5, color="black", fill="plum3")+
      labs(title = "Distribution of BMI among \npeople without pre/diabetes", x = "BMI") +
      theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
bmi0

grid.arrange(bmi1, bmi0, nrow = 1, ncol = 2)
```


**For Smoker**
```{r}
smoker1 <- ggplot(data = diabetes, aes(x = factor(Smoker), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Distribution of smokers", x = "smoking (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
smoker1
```

**For Stroke**

```{r}
stroke1 <- ggplot(data = diabetes, aes(x = factor(Stroke), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Stroke distribution", x = "have had a stroke (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
     scale_fill_manual(values=cbPalette)
    
stroke1
```

**For Heart Disease/Attack**

```{r}
heart_dis1 <- ggplot(data = diabetes, aes(x = factor(heart_dis), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Heart disease/attack distribution", x = "have heart disease/attack (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
heart_dis1
```

**For PhysActivity (if they've had any physical activity the past 30 days)**

```{r}
phys1 <- ggplot(data = diabetes, aes(x = factor(PhysActivity), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Physical activity distribution", x = "physical activity within 30 days (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
phys1
```
**For Fruits (daily consumption of fruits)**

```{r}
fruits1 <- ggplot(data = diabetes, aes(x = factor(Fruits), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Daily fruits consumption distribution", x = "had fruits (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
fruits1
```

**For Veggies**

```{r}
veg1 <- ggplot(data = diabetes, aes(x = factor(Veggies), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "Daily consumption of vegetables", x = "had vegetables (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
veg1
```

**For HvyAlcohol (adult men >= 14 drinks/wk, adult women >= 7 drinks/wk)**
```{r}
library(ggplot2)
hvyalc1 <- ggplot(data = diabetes, aes(x = factor(hvy_alc), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color = "black")+
      labs(title = "High alcohol consumption distribution", x = "high-alcohol consumption (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes")) +
      scale_fill_manual(values=cbPalette)
    
hvyalc1
```


**For AnyHealthcare**
```{r}
library(ggplot2)
any_health1 <- ggplot(data = diabetes, aes(x = factor(AnyHealthcare), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Health insurance availability", x = "have insurance (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
any_health1
```


**For NoDocbcCost (needed to see a doctor but did not due to cost in the past 12 months)**

```{r}
library(ggplot2)
nodoc1 <- ggplot(data = diabetes, aes(x = factor(NoDocbcCost), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Refusal of doctor visits due to cost", x = "visited doctor (0=yes, 1=no)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
nodoc1
```

**For GenHlth**


```{r}
genhlth1 <- ggplot(data = diabetes, aes(x = GenHlth, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of self-rated general health level (1 to 5)", x = "general health level") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
genhlth1
```

**For MentHlth (no. days of poor mental health, 1-30 days)**

```{r}
mentalh1 <- ggplot(data = diabts1, aes(x = MentHlth)) +
      geom_histogram(width = 0.5, color="black", fill="lightsteelblue")+
      labs(title = "Distribution of poor mental health days \namong people with pre/diabetes", x = "no. of days") +
      theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))
mentalh1

mentalh0 <- ggplot(data = diabts0, aes(x = MentHlth)) +
      geom_histogram(width = 0.5, color="black", fill="lightsteelblue")+
      labs(title = "Distribution of poor mental health days \namong people without pre/diabetes", x = "no. of days") +
      theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))
mentalh0

grid.arrange(mentalh1, mentalh0, nrow = 1, ncol = 2)
```

**For PhysHlth**

```{r}
physh1 <- ggplot(data = diabts1, aes(x = PhysHlth)) +
      geom_histogram(width = 0.5, color="black", fill="plum3")+
      labs(title = "Distribution of days with physical injuries \namong people with pre/diabetes", x = "no. of days") +
      theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))


physh0 <- ggplot(data = diabts0, aes(x = PhysHlth)) +
      geom_histogram(width = 0.5, color="black", fill="plum3")+
      labs(title = "Distribution of days with physical injuries \namong people without pre/diabetes", x = "no. of days") +
      theme(plot.title = element_text(size = 10, face = "bold", hjust = 0.5))


grid.arrange(physh1, physh0, nrow = 1, ncol = 2)
```

**For DiffWalk (if participant had serious difficulty walking/climbing stairs**

```{r}
diffwalk1 <- ggplot(data = diabetes, aes(x = factor(DiffWalk), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Difficulty walking/climbing stairs distribution", x = "have difficulty walking (0=no, 1=yes)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
diffwalk1
```

**For Sex**

```{r}
sex1 <- ggplot(data = diabetes, aes(x = factor(Sex), fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of sex", x = "sex (0 = female, 1 = male)") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
sex1
```

**For Age**
```{r}
age1 <- ggplot(data = diabetes, aes(x = Age, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of age groups", x = "age groups") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
age1
```

**For Education**

```{r}
edu1 <- ggplot(data = diabetes, aes(x = Education, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of education levels", x = "education level") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
edu1
```

**For Income**


```{r}
income1 <- ggplot(data = diabetes, aes(x = Income, fill=factor(diabetes))) +
      geom_bar(stat = "count", position=position_dodge(),width = 0.5, color="black")+
      labs(title = "Distribution of income levels", x = "income level") +
      theme(plot.title = element_text(face = "bold", hjust = 0.5))+
      guides(fill = guide_legend(title = "Diabetes"))+
      scale_fill_manual(values=cbPalette)
income1
```

Again, as the data is mostly categorical, we want to run chi-square tests to check for correlations.

```{r}
chisq.test(diabetes$diabetes, diabetes$HighBP)
chisq.test(diabetes$diabetes, diabetes$HighChol)
chisq.test(diabetes$diabetes, diabetes$CholCheck)
chisq.test(diabetes$diabetes, diabetes$Smoker)
chisq.test(diabetes$diabetes, diabetes$heart_dis)
chisq.test(diabetes$diabetes, diabetes$PhysActivity)
chisq.test(diabetes$diabetes, diabetes$Fruits)
chisq.test(diabetes$diabetes, diabetes$Veggies)
chisq.test(diabetes$diabetes, diabetes$hvy_alc)
chisq.test(diabetes$diabetes, diabetes$DiffWalk)
chisq.test(diabetes$diabetes, diabetes$Sex)
chisq.test(diabetes$diabetes, diabetes$AnyHealthcare)
chisq.test(diabetes$diabetes, diabetes$NoDocbcCost)
chisq.test(diabetes$diabetes, diabetes$GenHlth)
chisq.test(diabetes$diabetes, diabetes$MentHlth)
chisq.test(diabetes$diabetes, diabetes$Education)
chisq.test(diabetes$diabetes, diabetes$Income)
```

#### 4. Conclusion